---
title: "hw03-gapminder"
author: '@RetoSterchi'
date: "September 28, 2018"
output: github_document
---

# Homework 03: Use dplyr/ggplot2 to manipulate and explore data

## First step: Load packages

```{r message=FALSE, warning=FALSE}
require(gapminder)
require(ggplot2)
require(dplyr)
```


## Exploration 1: 
Get the maximum and minimum of GDP per capita for all continents.

```{r}
## Lets get the maximum GDP per capita, looked up under
## https://stackoverflow.com/questions/40157648/find-max-per-group-with-dplyr-in-r?rq=1
gapminder %>% 
  filter(year == 2007) %>% # only 2007
  group_by(continent) %>% # group by continent
  arrange(gdpPercap) %>% # arrange within continent
  slice(which.max(gdpPercap)) # take max per group
```

The same works with which.min()

```{r}
gapminder %>% 
  filter(year == 2007) %>% # only 2007
  group_by(continent) %>% # group by continent
  arrange(gdpPercap) %>% # arrange within continent
  slice(which.min(gdpPercap)) # take min per group
```

Or you can even just take the first row per group by specifiying slice(1)
```{r}
gapminder %>% 
  filter(year == 2007) %>% # only 2007
  group_by(continent) %>% # group by continent
  arrange(gdpPercap) %>% # arrange within continent
  slice(1) # take first row per group
```

## Exploration 2: 
Let's have a look at the spread of GDP per capita within the continents

```{r}
gapminder %>% 
  filter(year == 2007) %>% 
  group_by(continent) %>% 
  summarize(gdpMean = round(mean(gdpPercap),0),
            gdpMin = round(min(gdpPercap),0),
            gdpMax = round(max(gdpPercap),0),
            gdpSD = round(sd(gdpPercap),0))
  
```

This table already tells us a little bit about the data and the differences across the different continents. Some of them seem to have a larger spread than others (Americas, Asia, Europe).

```{r}
gapminder %>% 
  filter(year == 2007) %>% 
  group_by(continent) %>% 
  ggplot(aes(x = continent, y = gdpPercap, fill = continent)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.5) +
  geom_jitter(width = 0.2, aes(colour = continent), alpha = 0.8) +
  ggtitle("Spread of GDP per capita within the continents") +
  xlab("Continent") +
  ylab("GDP per capita (PPP Dollars)")
  
```

The boxplots confirms what we saw in the table. 


## Exploration 3: How is life expectancy changing over time on different continents?

Let's calculate the mean life expectancy per country first to get a sense of the data.

```{r}
gapminder %>% 
  group_by(continent, year) %>% 
  summarize(lifeExpMean = round(mean(lifeExp),0))
```

Now let's plot that to see how it developed over time (1952 - 2007). I think a plot showing the mean life expectancy per continent together with the life expectancy curves per country in the background could work nicely. Let's see.

```{r fig.height = 5, fig.width = 10}
gapminder %>% 
  group_by(continent, year) %>% 
  summarize(lifeExpMean = round(mean(lifeExp),0)) %>% # calculate mean per continent
  ggplot(aes(x = year, y = lifeExpMean, colour = continent)) +
  geom_line(data = gapminder, 
            aes(x = year, y = lifeExp, group = country), 
            alpha = 0.3) + # thinner line for each country
  geom_line(size = 1.2) + # thicker line for mean
  facet_grid(.~ continent) + # make plot for each continent
  theme(legend.position="none") + # we dont need a legend
  scale_x_continuous(breaks=seq(1950, 2010, 20)) + # specify ticks on x-axis
  ggtitle("Spread of GDP per capita within the continents") +
  xlab("Continent") +
  ylab("Life expectancy (Years)")

```

The difference per continent is tremendous! While Europe "started" with an average life expectancy of approximately 65 years in the 1950s, Africa and Asia's mean life expectancy was at approximately 40 to 45 years back then. Asia has caught up, though and its curve shows an impressive steepness.

Note to myself with regards to the graph: Think about the order of how the different layers get plotted (e.g., plot thin lines for the background first and then add the thicker lines in the foreground). Also, facet_grid() arranges the individual plots in one line, facet_wrap() just wraps the plots around the line (customizable though how many plots). In our case, it was more useul to arrange them all on the same line to make comparisons easier.

## Exploration 4: Can we find a good story with life expectancy in relation to income?
Let's see if we can do something with the life expectancy data and the income. Maybe let's start with a table where we summarize life expectancy (mean across countries) against the GPD per capita of individual countries (again, the mean across countries) for the most recent year in the dataset. For the sake of playing with different measures with can also think about including the weighted means (with regards to population). This might tell us something about disparities within a continent.

```{r}
gapminder %>% 
  filter(year == 2007) %>% 
  group_by(continent) %>% 
  summarize(lifeExpMean = round(mean(lifeExp), 1),
            gdpPercapMean = round(mean(gdpPercap),0),
            lifeExpMeanW = round(weighted.mean(lifeExp,pop),1),
            gdpPercapMeanW = round(weighted.mean(gdpPercap,pop),0))
```

```{r fig.height = 5, fig.width = 10}
gapminder %>%
  filter(year == 2007 & continent != "Oceania") %>% # exclude Oceania, since only two countries
  ggplot(aes(x = gdpPercap, y = lifeExp, group = country, colour = continent)) +
  geom_point() +
  xlab("GDP per capita (PPP dollars)") +
  ylab("Life expectancy (Years)") +
  ggtitle("Life expectancy vs. Income (2007, without Oceania)") +
  scale_fill_discrete(name="Continent") +
  scale_x_log10() # use a log scale
```

Looks like there is a relationship between these two parameters what makes sense if we consider what potentialy influences life expectancy (quality of sanitary facilities, child fatality, access to medical services etc.)

We can also see that there are seven countries in Africa where the life expectancy is above 70 years and comparable to many countries in Asia or some of Europe. Let's identify those seven countries with a query.

```{r}
gapminder %>%
  filter(year == 2007 & continent == "Africa" & lifeExp > 70)
```

Many of these countries seem to be former French colonies (https://en.wikipedia.org/wiki/List_of_French_possessions_and_colonies). Reunion actually is an overseas department of France. Is there potentially a historic component to this that the former French colonies are better of due to their ties to France?

What about African countries that have a fairly high GDP but still a lower life expectancy than our seven countries?

```{r}
gapminder %>%
  filter(year == 2007 & continent == "Africa" & lifeExp < 70 & gdpPercap > 9000)
```


Let's label these countries on the top of the graph with the help of the ggrepel package where we can specify the location of the text labels.

```{r fig.height = 5, fig.width = 10}
require(ggrepel)

  Query1 <- c("Reunion", "Libya", "Tunisia", "Mauritius", "Algeria", "Egypt", "Morocco")
  Query2 <- c("Botswana", "Equatorial Guinea", "Gabon", "South Africa")
  
  gapminder %>%
    filter(year == 2007 & continent != "Oceania") %>% # exclude Oceania, since only two countries
    ggplot(aes(x = gdpPercap, y = lifeExp, group = country, colour = continent, label = country)) +
    geom_point() +
    geom_text_repel(data = subset(gapminder, year == 2007 & country %in% Query1),
                    nudge_y = 90 - subset(gapminder, year == 2007 & 
                                            country %in% Query1)$lifeExp) + # this alignes the labels on the top
      geom_text_repel(data = subset(gapminder, year == 2007 & country %in% Query2),
                    nudge_x = 120 - subset(gapminder, year == 2007 & 
                                            country %in% Query2)$lifeExp) + # this alignes the labels on the right
    xlab("GDP per capita (PPP dollars)") +
    ylab("Life expectancy (Years)") +
    ggtitle("Life expectancy vs. Income (2007, without Oceania)") +
    scale_fill_discrete(name="Continent") +
    scale_x_log10() # use a log scale
```

Two of these countries are again former French colonies. Botswana is a former British colony and South Africa was both a Dutch and later a British colony. So there only just a little support for the theory that the former French colonies might be better of. Could it be because they are geographically closer to Europe? Most of the top seven African countries are in the Mediterranian region... I guess we would need additional data to support more of this exploration.


## Let's explore some other ways to make pleasant tables

Let's check out the 'pander' package. This seems to be a nice way of showing tables in a report. Queries can directly by passed on to the pander() function and are displayed in a pleasing way instead of the tibble output format.

Check out this link for more details on how to customize the output: http://rapporter.github.io/pander/

```{r message=FALSE, warning=FALSE}
# install.packages('pander')
require(pander)

## This makes nice table for a report
pander(gapminder %>%
  filter(year == 2007 & continent == "Africa" & lifeExp > 70))

## This makes a simple table in a console style output
pandoc.table(gapminder %>%
  filter(year == 2007 & continent == "Africa" & lifeExp > 70),
  style = "simple")

```
```{r}
## We can also add captions to such tables and specify the alignment
pander(gapminder %>%
  filter(year == 2007 & continent == "Africa" & lifeExp > 70),
  caption = "Hello, this is a table caption!",
  justify = c('left', 'center', 'center', 'center', 'right', 'right'))
```

